Mechanisation of Session-based Lambda Calculus Type Systems

### 2014-12-05 proposal/proposal.tex
Finished 3rd (and possibly final) draft.

\subsubsection{Preamble}

@@

They represent dual intuitionistic linear logic as a sequent calculus
providing logical connectives which have session-based interpretations in
$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL:
\input{sill-branch-choice-rules}

Note that branch and choice both have some notion of branching (R rules for
$\oplus$/L rules for $\with$) and at least for \& this is counter-intuitive in
a session-based setting, since the intutition for \& is that of
\emph{offering} a choice between alternatives rather than \emph{making} the
choice. Here, however, the intuition should be extended to account for its
location in the typing judgement. The channel is on the left of the sequent,
and \citeauthor{Caires:2010:STI} suggest one consider the context as services
being provided to the typed process. Thus, the type of channels in the context
is really the type of the other endpoint of the channel, and the typed process
behaves according to the dual protocol. For example, considering
$\SILLbranchLTwo$, and taking $A^\bot$ to represent the dual of a protocol $A$
as defined by \citeauthor{Wadler:2014}~\cite{Wadler:2014}, the typed process
is behaving according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting
from alternatives (in this case, $\inrterm$) and then behaving according to
the selected protocol (in this case, protocol $B^\bot$).

@@

Note, I am using $\tensor{}{}$ for session output and $\lfn{}{}$ for session
input following the presentation by \citeauthor{Caires:2010:STI}. In the
protocol defined above, a client selects between buying a book or receiving a
price quote for a book. In the buy option case, the client sends the name of
the book to the server and their card details, then the server sends the
receipt for the purchase. In the quote option case, the client sends the name
of the book and receives a price quote from the server. I have replaced the
types $N$ and $I$ used by \citeauthor{Caires:2010:STI} with more meaningful
type names $Book$, $Card$, $Quote$ and $Receipt$ where appropriate; all are
synonyms for the type $\1$ which denotes the terminal session type. The client
and server process bodies are as follows:
\begin{gather*}
\sdef{ClientBody_c}
     {\CPinr{c}{\SLoutput{1984}{c}{\SLinput{c}{quote}{\0}}}}\\
\sdef{ServerBody_c}
     {\CPcase{c}
             {\SLinput{c}{book}{\SLinput{c}{card}{\SLoutput{receipt}{c}{\0}}}}
             {\SLinput{c}{book}{\SLoutput{quote}{c}{\0}}}}
\end{gather*}

@@

\citeauthor{Wadler:2014}'s recent work~\cite{Wadler:2014} brings together a
process calculus, CP, based on the earlier work by
\citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by
\citeauthor{Gay:2010:LAST}. The work describes a continuation-passing style
(CPS) translation from GV to CP treating session types as propositions in a
classical linear logic. The process calculus is slightly different to that
described by $\pi$DILL~\cite{Caires:2010:STI}, apart from being classical in
nature, the constructs for server replication are modified; CP provides a
weakening and contraction rules whereas these are encoded in one rule for
process reduction by \citeauthor{Caires:2010:STI}. Additionally, rather than a
two-sided sequent calculus, \citeauthor{Wadler:2014} uses one-sided sequents
for defining the process calculus, which provides a more intuitive
presentation of session type duality and clearer correspondence to a
$\lambda$-calculi surface syntax. As presented, GV does not support all
features of CP and thus the translation is one way (GV to
CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to
provide support for polymorphism and replication, following a similar CPS
translation scheme. Lastly,
\citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational
semantics for GV and their system has greater similarities with LAST than
prior GV works, for example, by treating send and receive as linear functions
rather than language primitives.

%% Wadler also dispenses with the structural rules; subsumed by the Assoc and
%% Swap typing derivations and the cut reduction (double check why he does
%% this)

@@

While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
may need to alter the approach slightly to avoid tricky formulations. For
instance, instead of providing $n$-ary branch and choice in GV, I provide
only binary versions of these operators to simplify the development. This
change is not restrictive however, since CP itself provides only binary
versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
technical note, I amalgamate all types into one inductive definition and
provide a predicate for restricting typing rules to consider only valid
session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
recursive definitions; session types are types which may contain types as
subcomponents (arguments to send, for example). Unfortunately, handling
mutually inductive definitions in Coq can be quite involved; requiring one to
either rely on the Coq system to provide a stronger mutual induction principle
or defining one manually. Such a definition is possible, but it complicates
elimination via the induction principle.
% ( since the standard induction tactic will not work?)

@@

\begin{tabularx}{\linewidth}{|c|X|}\hline
Deliverable \# &  Description \\ \hline
\Deliverable{1}{Mechanisation of Wadler's system}
\DSubD{1}{*}{Mechanisation of Wadler's system}
\Deliverable{2}{Closure side-effecting extension}
\DSubD{2}{1}{Implementation of the GV operational semantics}
\DSubD{2}{2}{Implementation of the closure side-effecting extension}
\Deliverable{3}{Report write-up}
\DSubD{3}{1}{Description of approach, Wadler mechanisation and related work}
\end{tabularx}

@@

\begin{center}
\begin{tabular}{|*{5}{c|}}\hline
\multicolumn{5}{|c|}{\textbf{Estimated Work Plan}}\\ \hline
Task & Start Date & Duration (days) & End Date & Deliverables \\ \hline
\DTask{1}{1}{8th December}{15}{23rd December}{-}
\DTask{3}{1}{20th December}{3}{23rd December}{-}
\DTask{1}{2}{24th December}{17}{10th January}{D1}
\DTask{3}{2}{7th January}{11}{18th January}{D3.1}
\DTask{2}{1}{11th January}{7}{18th January}{-}
\DTask{2}{2}{19th January}{18}{6th Feburary}{D2.1}
%\DTask{3}{3}{
\end{tabular}
\end{center}

@@

For the remainder, I shall look to extend the definitions of GV and CP using
the definitions I have provided. As discussed previously, it would be
interesting to study a system for supporting side-effecting functions (aka
closures) for session channels. In order to achieve this, operational
semantics for GV would need to be defined and the recent draft by
\citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
do this~\cite{Lindley:2014:SPS}. The task allocation time indicated in the
work plan table. \IDTsk{2}{1} is a theoretical investigation into the
operational semantics, deciding whether the system by
\citeauthor{Lindley:2014:SPS} offers a suitable basis for the extension, and
investigating the details of the side-effecting closure extension (assuming an
appropriate base operational semantics for GV). \IDTsk{2}{2} is then the
implementation of the suitable operational semantics (without side-effecting
closure), culminating in the first part of the second deliverable
(D2.1). Lastly, \IDTsk{2}{3} is an allocation of time to the task of
implementing the extension permitting closures with side-effects. In my work
plan, I have assumed \IDTsk{2}{2} implemented an operational semantics that
eases this particular extension which justifies the task allocation time. Both
\IDTsk{2}{2} and \IDTsk{2}{3} may also include revisions to the theoretical
treatment started in \IDTsk{2}{1} as a consequence of initial design
oversights.

There is a risk of being unable to perform the side-effecting extension as
sketched in \S \ref{sec:approach}. I shall review and update the above plan
throughout the project as I proceed to the various milestones I have set out,
and if required will alter the above plans if necessary.

>>>>

\subsubsection{Preamble}\label{sec:pre}

@@

They represent dual intuitionistic linear logic as a sequent calculus
providing logical connectives which have session-based interpretations in
$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL, which have
only two alternatives (cf. multiple alternatives in the standard definition in
\S~\ref{sec:pre}):
\input{sill-branch-choice-rules}

Note that branch and choice both have some notion of branching (R rules for
$\oplus$/L rules for $\with$) and at least for \& this is counter-intuitive in
a session-based setting, since the intutition for \& is that of
\emph{offering} (receiving) a choice between alternatives rather than
\emph{making} (sending) the choice. Here, however, the intuition should be
extended to account for its location in the typing judgement. The channel is
on the left of the sequent, and \citeauthor{Caires:2010:STI} suggest one
consider the context as services being provided to the typed process. Thus,
the type of channels in the context is really the type of the other endpoint
of the channel, and the typed process behaves according to the dual
protocol. For example, considering $\SILLbranchLTwo$, and taking $A^\bot$ to
represent the dual of a protocol $A$ as defined by
\citeauthor{Wadler:2014}~\cite{Wadler:2014}, the typed process is behaving
according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting from
alternatives (in this case, $\inrterm$) and then behaving according to the
selected protocol (in this case, protocol $B^\bot$).

@@

Note, I am using $\tensor{}{}$ for session output and $\lfn{}{}$ for session
input following the presentation by \citeauthor{Caires:2010:STI}. In the
protocol defined above, a client selects between buying a book or receiving a
price quote for a book. In the buy option case, the client sends the name of
the book to the server and their card details, then the server sends the
receipt for the purchase. In the quote option case, the client sends the name
of the book and receives a price quote from the server. I have replaced the
types $N$ and $I$ used by \citeauthor{Caires:2010:STI} with more meaningful
type names $Book$, $Card$, $Quote$ and $Receipt$ where appropriate; all are
synonyms for the type $\1$ which denotes the terminal session type in this
system. The client and server process bodies are as follows:
\begin{gather*}
\sdef{ClientBody_c}
     {\CPinr{c}{\SLoutput{1984}{c}{\SLinput{c}{quote}{\0}}}}\\
\sdef{ServerBody_c}
     {\CPcase{c}
             {\SLinput{c}{book}{\SLinput{c}{card}{\SLoutput{receipt}{c}{\0}}}}
             {\SLinput{c}{book}{\SLoutput{quote}{c}{\0}}}}
\end{gather*}

@@

\citeauthor{Wadler:2014}'s recent work~\cite{Wadler:2014} brings together a
process calculus, CP, based on the earlier work by
\citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by
\citeauthor{Gay:2010:LAST}. GV presents session types using the standard
presentation as described in \S~\ref{sec:pre}. CP presents linear logical
connectives with session type interpretations inspired by
\citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI}. \citeauthor{Wadler:2014}
describes a continuation-passing style (CPS) translation from GV session types
to CP classical linear logic propositions. The process calculus CP is slightly
different to $\pi$DILL by \citeauthor{Caires:2010:STI}, apart from being
classical in nature, the constructs for server replication are modified; CP
provides a weakening and contraction rules whereas these are encoded in one
rule for process reduction by \citeauthor{Caires:2010:STI}. Additionally,
rather than a two-sided sequent calculus, \citeauthor{Wadler:2014} uses
one-sided sequents for defining the process calculus, which provides a more
intuitive presentation of session type duality and clearer correspondence to a
$\lambda$-calculi surface syntax, simplifying the translation presentation. As
presented, GV does not support all features of CP and thus the translation is
one way (GV to CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP}
extend GV to provide support for polymorphism and replication, following a
similar CPS translation scheme. Lastly,
\citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational
semantics for GV and their system has greater similarities with LAST than
prior GV works, for example, by treating send and receive as linear functions
rather than language primitives.

@@

While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
may need to alter the approach slightly to avoid tricky formulations. For
instance, instead of providing $n$-ary branch and choice in GV, I provide
only binary versions of these operators to simplify the development. This
change is not restrictive however, since CP itself provides only binary
versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
technical note, I amalgamate all types into one inductive definition and
provide a predicate for restricting typing rules to consider only valid
session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
recursive definitions; session types are types which may contain types as
subcomponents (arguments to send, for example). Unfortunately, handling
mutually inductive definitions in Coq can be quite involved; requiring one to
either rely on the Coq system to provide a stronger mutual induction principle
or defining one manually. Such a definition is possible, but it complicates
elimination via the induction principle.

@@

\begin{tabularx}{\linewidth}{|c|X|}\hline
Deliverable \# &  Description \\ \hline
\Deliverable{1}{Mechanisation of Wadler's system}
D1 & Mechanisation of Wadler's system \\ \hline
\Deliverable{2}{Closure side-effecting extension}
\DSubD{2}{1}{Implementation of the GV operational semantics}
\DSubD{2}{2}{Implementation of the side-effecting closure extension}
\Deliverable{3}{Report write-up}
\DSubD{3}{1}{Description of approach, Wadler mechanisation and related work}
\DSubD{3}{2}
      {Description of GV operational semantics (w/o side-effecting closure)}
\DSubD{3}{3}{Description of side-effecting closure and final version}
\end{tabularx}

@@

\begin{center}
\begin{tabular}{|*{5}{c|}}\hline
\multicolumn{5}{|c|}{\textbf{Estimated Work Plan}}\\ \hline
Task & Start Date & Duration (days) & End Date & Deliverables \\ \hline
\DTask{1}{1}{8th December}{15}{23rd December}{-}
\DTask{3}{1}{20th December}{3}{23rd December}{-}
\DTask{1}{2}{24th December}{17}{10th January}{D1}
\DTask{3}{2}{7th January}{11}{18th January}{D3.1}
\DTask{2}{1}{11th January}{14}{25th January}{-}
\DTask{2}{2}{23th January}{21}{16th Feburary}{D2.1}
\DTask{3}{3}{13th Feburary}{10}{23rd Feburary}{D3.2}
\DTask{2}{3}{17th Feburary}{14}{3rd March}{D2.2}
\DTask{3}{4}{1st March}{14}{15th March}{D3.3}
\DTask{*}{*}{16th March}{39}{24th April}{-}
\end{tabular}
\end{center}

@@

For the remainder, I shall look to extend the definitions of GV and CP using
the definitions I have provided. As discussed previously, it would be
interesting to study a system for supporting side-effecting functions (aka
closures) for session channels. In order to achieve this, operational
semantics for GV would need to be defined and the recent draft by
\citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
do this~\cite{Lindley:2014:SPS}. \IDTsk{2}{1} is a theoretical investigation
into the operational semantics, deciding whether the system by
\citeauthor{Lindley:2014:SPS} offers a suitable basis for the extension, and
investigating the details of the side-effecting closure extension (assuming an
appropriate base operational semantics for GV). \IDTsk{2}{2} is then the
implementation of the suitable operational semantics (without side-effecting
closure), culminating in the first part of the second deliverable
(D2.1). Following this implementation work, I plan to write up the operational
semantics in the report (\IDTsk{3}{3}) which constitutes the second draft of
the report deliverable (D3.2). Lastly, \IDTsk{2}{3} is an allocation of time
to the task of implementing the extension permitting closures with
side-effects which includes necessary changes to the typing judgements and
translation between GV and CP. In my work plan, I have assumed \IDTsk{2}{2}
implemented an operational semantics that eases this particular extension
which justifies the task allocation time. Both \IDTsk{2}{2} and \IDTsk{2}{3}
may also include revisions to the theoretical treatment started in
\IDTsk{2}{1} as a consequence of initial design oversights. \IDTsk{3}{4}
constitutes the final version of the report, including discussion of the
side-effecting closure extension.

There is a risk of being unable to perform the side-effecting extension as
sketched in \S \ref{sec:approach}. Additionally, there is considerable slack
time identified in the work plan, indicated by \IDTsk{*}{*}. I believe these
task allocation time estimates to be an accurate reflection of the work
involved. However, the feasibility study is necessarily limited in scope, so I
shall review and update the above plan throughout the project as I proceed to
the various milestones I have set out, and if required will alter the above
plans if necessary.
